HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-27 scnn.tex
small modification in result section

\todo{learned histogram \citep{wang2016learnhist} if it shows good results}

\todo{stats of diffused maps \citep{khasanova2017tigranet}. Provides stats at multiple scales?}

\todo{perf w.r.t. different functions $f$. Here or in the experiment section?}

@@

We make full sky N-body simulations for two parameter sets for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ \todo{(MORE PARAMS HERE - CHECK WITH RAPHAEL)}.
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in \figref{psd_sigma3}. In the pre-processing step, we also remove the mean of each map.

@@

For all the experiments, we work with centered Gaussian noise.
While the noise model is real mass maps often has a slightly different distribution, Gaussian should be a sufficient model to demonstrate the performance of our method.
To improve the stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between $50$ and $290$ epochs depending on the complexity and the size of
the problem. $20\%$ of the training set that we call "validation set" was left out of the training process and simply used to assess the global performance and overfitting of
the network.
Training took less than a day for each model using an NVIDIA Tesla K20.

We compare the performance of the spherical CNN against two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities (PSD) and b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified use a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better results
while having worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in Appendix~\ref{sec:dataset_augmentation}. We tuned the parameter using cross-validation with the validation set.

@@

We present in \figref{results} the results for 5 different noise
levels and $3$ different sample areas. The spherical CNN has a significant lead for all levels of noise and sample areas.
In particular, the difference seems to increase as the problem becomes more complex, i.e., as the order is increased and there is less information in each sample.
This is likely due to the fact that the SCNN can learn optimal features for the problem and do not rely on our humanly limited feature design.

We observe that the PSD features, i.e.,
one of the most used techniques for model discrimination in cosmology, is unable
to perform the desired task. The main reason is that the two different models
have very similar PSDs (see \figref{psd_sigma3}). As a result, one
needs to use other statistics to solve the problem. In this case, the
histogram features that have been built for the problem allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the
distribution of pixels, they do not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information.
Note that without this graph structure, a fully connected neural network approach would have to be used. Problematically, fully connected neural networks do not scale well with the number of pixels.

>>>>

Other statistical layers are possible. In particular, we also implemented a learned histogram from \citep{wang2016learnhist}. Additionaly \citep{khasanova2017tigranet} propose to apply differen diffusion on the final maps to obtain statistics at different sclaes.

@@

We make full sky N-body simulations for two parameter sets for $\Lambda \rm{CDM} $ cosmological model: model 1 ($\Omega_m=0.31, \sigma_8=0.82$) and model 2 ($\Omega_m=0.26, \sigma_8=0.91$), where $\Omega_m$ is the matter density in the universe and $\sigma_8$ is the normalisation of the matter power spectrum.
Other parameter parameters were set to $h=0.7$ \todo{(MORE PARAMS HERE - CHECK WITH RAPHAEL)}.
These parameters were chosen to have the same spherical harmonic power spectrum for $\ell<1000$.
That means that it is not possible to distinguish between these cosmological models using these maps if the this range of scales is used.
We found that for $\ell>1000$ the differences in power spectrum is $~5\%$.
To remove this information, we additionally smooth the spherical maps with a Gaussian kernel of radius $3$ arcmins. The resulting PSD are displayed in \figref{psd_sigma3}. In the pre-processing step, we also remove the mean of each map and downsample them to a resolution of of $Nside=1024$.

@@

For all the experiments, we work with centered Gaussian noise.
While the noise model of real mass maps often has a slightly different distribution, Gaussian should be a sufficient model to demonstrate the performance of our method.
To improve the stability of the spherical CNN, we slowly increase the amount of noise added to
the training data over the iterations. Furthermore in order to avoid
overfitting, the noise is randomly produced during the training process. We trained our
model between $50$ and $290$ epochs depending on the complexity and the size of
the problem. $20\%$ of the training set that we call "validation set" was left out of the training process and simply used to assess the global performance and overfitting of
the network.
Training took less than a day for each model using an NVIDIA Tesla K20.

We compare the performance of the spherical CNN against two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities (PSD) and b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified using a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better results
while having significanly worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in Appendix~\ref{sec:dataset_augmentation}. We tuned the regularization parameter by cross-validating with the validation set.

@@

We present in \figref{results} the results for $5$ different noise
levels and $3$ different sample areas. The spherical CNN has a significant lead for all levels of noise and sample areas.
In particular, the difference seems to increase as the problem becomes more complex, i.e., as the order is increased and there is less information in each sample.
This is likely due to the fact that the SCNN can learn optimal features for the problem and do not rely on our humanly limited feature design.

We observe that the PSD features, i.e.,
one of the most used techniques for model discrimination in cosmology, is unable
to perform the desired task. The main reason is that the two different models
have very similar PSDs (see \figref{psd_sigma3}). As a result, one
needs to use other statistics to solve the problem. In this case, the
histogram features allow for a significant improvement over the PSD ones.
Unfortunately, while the histogram features contain information about the
distribution of pixels, they do not include any spatial information. In
contrast, both the PSD features and the SCNN can exploit the spatial
information.
Note that without this graph structure, a fully connected neural network approach would have to be used. Problematically, fully connected neural networks do not scale well with the number of pixels.

We have many different spherical neural network architectures and made a few empirical observations. 
First Chebyshef polynomial seems more stable than monomials. Nevertheless once the network is trained, we could not observe a significant accuracy difference.
Second, using a statistical layer improved slightly the result for our specific problem. We tried $4$ different layers `mean', `variance', `mean + variance' and `histogram'. All but the `variance' were providing the same range of accurancy, the variance being slightly worse.
Third, dropout did not help in the precence of noise. This is probably du to the fact that the dataset was augmented indefinitely
Fourth, especially for difficult problem (order 2 or 4, and the presence of large noise) having a large architecture (many filters) and doing many iterations were necessary to achieve the result presented.
